On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-11-25 paper.tex
Rewriting of Section 2; math expressions not checked

The optimal pairs in this specific case have $j=i$ or $j=i-1$, as seen in the figure. Due to symmetry, any pair $(i,i-1)$ could be replaced by $(i-1,i)$. This illustrates that the set of optimal pairs is not unique in general.

@@

\begin{proof}
For each $\tiles$, the pair $(i_\tiles,j_\tiles)$ should be chosen as that on the line $i+j-1=\tiles$ which minimizes $(i-2)^2 a^2 + (j-2)^2 b^2$. This allows maximizing the sum $i+j-1$, and thus the number of visited tiles, for a given length restriction (direct problem); or minimizing the required lengths for a specified number of visited files (inverse problem).

Consider, for the moment, $i$, $j$ as if they were real-valued variables, and denote $x=ia$, $y=jb$. The line $i+j-1 = \tiles$ then becomes
\begin{equation}
\label{eq: line 1}
\frac x a + \frac y b = \tiles + 1,
\end{equation}
and $(i-2)^2 a^2 + (j-2)^2 b^2$ is expressed as $(x-2a)^2+(y-2b)^2$. The point minimizing this quadratic function along \eqref{eq: line 1} is the intersection of that line with the perpendicular line passing through $(2a,2b)$, that is,
\begin{equation}
\label{eq: line 2}
y = \frac {a (x-2a)} b + 2b.
\end{equation}
This is depicted for $a=1.35$, $b=1$, $\tiles=7$ in Figure~\ref{fig: min suff set, form}, where \eqref{eq: line 2} is shown with solid line. Solving the system of equations \eqref{eq: line 1} and \eqref{eq: line 2} gives 
\begin{align}
\frac x a - 2 &= \frac{(t-3)b^2}{a^2+b^2}, \\
\frac y b - 2 &= \frac{(t-3)a^2}{a^2+b^2}.
\end{align}
In terms of the real-valued variables $i$, $j$, the solution $(\isolr, \jsolr)$ is
\begin{align}
\label{eq: prop: min suff set, form: isolr}
\isolr &= \frac{(t-3)b^2}{a^2+b^2} + 2, \\
\label{eq: prop: min suff set, form: jsolr}
\jsolr &= \frac{(t-3)a^2}{a^2+b^2} + 2.
\end{align}
The corresponding point $(\isolr a, \jsolr b)$ is shown in Figure~\ref{fig: min suff set, form} with a square marker.

@@

The variables $i$, $j$ are actually limited to integer values. The pair $(i_\tiles, j_\tiles)$ that minimizes $(i-2)^2 a^2 + (j-2)^2 b^2$ along the line $i+j-1=\tiles$ restricted to integer values is either $(\lceil \isolr \rceil, \lfloor \jsolr \rfloor)$ or $(\lfloor \isolr \rfloor, \lceil \jsolr \rceil)$, whichever gives $(i_\tiles a, j_\tiles b)$ closest to $(\isolr a, \jsolr b)$, as illustrated in Figure~\ref{fig: min suff set, form}. In case of a tie the first of the two pairs is chosen. This can be expressed as
\begin{align}
\label{eq: prop: min suff set, form: isoli}
i_\tiles &= \left\lfloor \isolr + \frac 1 2 \right\rfloor, \\
\label{eq: prop: min suff set, form: jsoli}
j_\tiles &= \left\lceil \jsolr - \frac 1 2 \right\rceil,
\end{align}
which corresponds to rounding $\isolr$ and $\jsolr$ to the closest integers, with ties resolved in opposite directions. Combining \eqref{eq: prop: min suff set, form: isolr}--\eqref{eq: prop: min suff set, form: jsoli}
% *!*
yields \eqref{eq: min suff set, form, i} and \eqref{eq: min suff set, form, j}.

From \eqref{eq: prop: min suff set, form: isolr} and \eqref{eq: prop: min suff set, form: isoli},
\begin{equation}
\label{eq: prop: min suff set, form: isoli ineq}
\frac{(t-3)b^2}{a^2+b^2} + \frac 3 2 < i_\tiles \leq \frac{(t-3)b^2}{a^2+b^2} + \frac 5 2,
\end{equation}
and similarly, from \eqref{eq: prop: min suff set, form: jsolr} and \eqref{eq: prop: min suff set, form: jsoli},
\begin{equation}
\label{eq: prop: min suff set, form: jsoli ineq}
\frac{(t-3)a^2}{a^2+b^2} + \frac 3 2 \leq j_\tiles < \frac{(t-3)a^2}{a^2+b^2} + \frac 5 2.
\end{equation}
Considering the first inequality in \eqref{eq: prop: min suff set, form: isoli ineq} and the second in \eqref{eq: prop: min suff set, form: jsoli ineq} as equalities and eliminating $\tiles$ gives \eqref{eq: upper bound, line}. The pair $(i_\tiles, j_\tiles)$ is strictly below the line \eqref{eq: upper bound, line} because the used inequalities are strict. Similarly, the bounding line \eqref{eq: lower bound, line} results from the second inequality in \eqref{eq: prop: min suff set, form: isoli ineq} and the first in \eqref{eq: prop: min suff set, form: jsoli ineq}, and the fact that both inequalities are not strict implies that the bound \eqref{eq: lower bound, line} can actually be attained.
\end{proof}

@@

For $a^2/b^2$ arbitrary, the number of pairs in $\mss$ with the same $i$ coordinate, or with the same $j$, is in general irregular, because the lines \eqref{eq: upper bound, line} and \eqref{eq: lower bound, line} do not follow a ``natural'' direction of the grid. This happens for instance in Figure~\ref{fig: pairs_bounds_1p35}, where the number of pairs for each $i$ equals either $2$ or $3$ without a clear pattern.\footnote{Strictly, there is a periodic pattern whenever $a^2/b^2$ is rational, which is the case in Figure~\ref{fig: pairs_bounds_1p35}. However, it is not easily discernible unless $a^2/b^2$ is a ratio of small numbers.}
On the other hand, a simple pattern arises when $a^2/b^2$ or $b^2/a^2$ is a natural number, as seen in Figures~\ref{fig: pairs_bounds_1} and \ref{fig: pairs_bounds_sqrt2}.

>>>>

The optimal pairs in this specific case have the form $(i,i)$ or $(i,i-1)$, as seen in the figure. Due to symmetry, any pair $(i,i-1)$ could be replaced by $(i-1,i)$. This illustrates that the set of optimal pairs is not unique in general.

@@

\begin{proof}
For each $\tiles$, the pair $(i_\tiles,j_\tiles)$ should be chosen as that on the line $i+j-1=\tiles$ which minimizes $(i-2)^2 a^2 + (j-2)^2 b^2$. This allows maximizing the sum $i+j-1$, and thus the number of visited tiles, for a given length restriction; or visiting a specified number of files with lengths as small as possible.

Consider, for the moment, $i$, $j$ as if they were real-valued variables, and denote $x=ia$, $y=jb$. The line $i+j-1 = \tiles$ then becomes
\begin{equation}
\label{eq: line 1}
\frac x a + \frac y b = \tiles + 1,
\end{equation}
and $(i-2)^2 a^2 + (j-2)^2 b^2$ is expressed as $(x-2a)^2+(y-2b)^2$. The point minimizing this quadratic function along the line \eqref{eq: line 1} is the intersection of that with the perpendicular line passing through $(2a,2b)$,
\begin{equation}
\label{eq: line 2}
y = \frac {a (x-2a)} b + 2b.
\end{equation}
An example with $a=1.35$, $b=1$, $\tiles=7$ is depicted in Figure~\ref{fig: min suff set, form}, where \eqref{eq: line 2} is shown with solid line. Solving the system of equations \eqref{eq: line 1} and \eqref{eq: line 2} gives 
\begin{align}
\frac x a - 2 &= \frac{(t-3)b^2}{a^2+b^2}, \\
\frac y b - 2 &= \frac{(t-3)a^2}{a^2+b^2}.
\end{align}
In terms of the real-valued variables $i$, $j$, the solution $(\isolr, \jsolr)$ is thus
\begin{align}
\label{eq: prop: min suff set, form: isolr}
\isolr &= \frac{(t-3)b^2}{a^2+b^2} + 2, \\
\label{eq: prop: min suff set, form: jsolr}
\jsolr &= \frac{(t-3)a^2}{a^2+b^2} + 2.
\end{align}
The corresponding point $(\isolr a, \jsolr b)$ is shown in Figure~\ref{fig: min suff set, form} with a square marker.

@@

The variables $i$, $j$ are actually limited to integer values. The pair $(i_\tiles, j_\tiles)$ that minimizes $(i-2)^2 a^2 + (j-2)^2 b^2$ along the line $i+j-1=\tiles$ restricted to integer values is either $(\lceil \isolr \rceil, \lfloor \jsolr \rfloor)$ or $(\lfloor \isolr \rfloor, \lceil \jsolr \rceil)$, whichever gives $(i_\tiles a, j_\tiles b)$ closest to $(\isolr a, \jsolr b)$, as illustrated in Figure~\ref{fig: min suff set, form}. In case of a tie the first of the two pairs is (arbitrarily) chosen. This can be expressed as
\begin{align}
\label{eq: prop: min suff set, form: isoli}
i_\tiles &= \left\lfloor \isolr + \frac 1 2 \right\rfloor, \\
\label{eq: prop: min suff set, form: jsoli}
j_\tiles &= \left\lceil \jsolr - \frac 1 2 \right\rceil,
\end{align}
which corresponds to rounding $\isolr$ and $\jsolr$ to the closest integers, with ties resolved in opposite directions. Combining \eqref{eq: prop: min suff set, form: isolr}--\eqref{eq: prop: min suff set, form: jsoli}
% *!*
yields \eqref{eq: min suff set, form, i} and \eqref{eq: min suff set, form, j}.

From \eqref{eq: prop: min suff set, form: isolr} and \eqref{eq: prop: min suff set, form: isoli},
\begin{equation}
\label{eq: prop: min suff set, form: isoli ineq}
\frac{(t-3)b^2}{a^2+b^2} + \frac 3 2 < i_\tiles \leq \frac{(t-3)b^2}{a^2+b^2} + \frac 5 2,
\end{equation}
and similarly, from \eqref{eq: prop: min suff set, form: jsolr} and \eqref{eq: prop: min suff set, form: jsoli},
\begin{equation}
\label{eq: prop: min suff set, form: jsoli ineq}
\frac{(t-3)a^2}{a^2+b^2} + \frac 3 2 \leq j_\tiles < \frac{(t-3)a^2}{a^2+b^2} + \frac 5 2.
\end{equation}
Considering the first inequality in \eqref{eq: prop: min suff set, form: isoli ineq} and the second in \eqref{eq: prop: min suff set, form: jsoli ineq} as equalities and eliminating $\tiles$ gives \eqref{eq: upper bound, line}. The pair $(i_\tiles, j_\tiles)$ is strictly below the line \eqref{eq: upper bound, line} because the used inequalities are strict. Similarly, the bounding line \eqref{eq: lower bound, line} results from the second inequality in \eqref{eq: prop: min suff set, form: isoli ineq} and the first in \eqref{eq: prop: min suff set, form: jsoli ineq}, and the fact that those inequalities are not strict implies that the bound \eqref{eq: lower bound, line} can actually be attained.
\end{proof}

@@

For $a^2/b^2$ arbitrary, the number of pairs in $\mss$ with the same $i$ coordinate, or with the same $j$, is in general irregular, because the lines \eqref{eq: upper bound, line} and \eqref{eq: lower bound, line} do not follow a ``natural'' direction of the grid. This happens for instance in Figure~\ref{fig: pairs_bounds_1p35}, where the number of pairs for each $i$ equals either $2$ or $3$ without a clear pattern.\footnote{Strictly, there is a periodic pattern whenever $a^2/b^2$ is rational, which is the case in Figure~\ref{fig: pairs_bounds_1p35}. However, the pattern is not easily discernible unless $a^2/b^2$ is a ratio of small numbers.}
On the other hand, a simple pattern arises when $a^2/b^2$ or $b^2/a^2$ is a natural number, as seen in Figures~\ref{fig: pairs_bounds_1} and \ref{fig: pairs_bounds_sqrt2}.
